---
title: "FFP Activities"
author: "Laura Hughes"
date: "February 18, 2016"
output: word_document
---

```{r setup, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
library(stringr)
```

## Activities for 2015 reporting. 
```{r import, echo = FALSE, cache = TRUE}
activ = read.csv('~/Documents/USAID/Kyla - FFP reshaping/Development_activities_2016-02-19.csv') %>% 
    filter(pgrmType == 'Development',
           isActive == 1, 
           ARRyear == 2015)

allActiv = read.csv('~/Documents/USAID/Kyla - FFP reshaping/allFFP_activities_2016-02-19.csv') %>% 
    filter(pgrmType == 'Development',
           isActive == 1, 
           ARRyear == 2015)
```

### MCH : number of projects per country
** Note: contains both 1st or 2nd category labeled as MCH.  Similar for all below**


```{r MCH_country, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Maternal') | 
                         str_detect(category2, 'Maternal')) %>% 
               filter(ARRyear==2015) %>% 
        group_by(Country) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```

### MCH : breakdown of subcategories for each country
```{r MCH_subcat, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Maternal') | 
                         str_detect(category2, 'Maternal')) %>% 
        group_by(Country, subcategory) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```


### MCH : breakdown of activities, overall
```{r MCH_activ, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Maternal') | 
                         str_detect(category2, 'Maternal')) %>% 
        group_by(activity) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```



### Ag : number of projects per country
```{r Ag_country, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Agriculture') | 
                         str_detect(category2, 'Agriculture')) %>% 
        group_by(Country) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```

### Ag : breakdown of subcategories for each country
```{r Ag_subcat, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Agriculture') | 
                         str_detect(category2, 'Agriculture')) %>% 
        group_by(Country, subcategory) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```


### Ag : breakdown of activities, overall
```{r Ag_activ, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'Agriculture') | 
                         str_detect(category2, 'Agriculture')) %>% 
        group_by(activity) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```


### WASH : number of projects per country
```{r wash_country, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(Country) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```

### WASH : breakdown of subcategories for each country
```{r wash_subcat, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(Country, subcategory) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```


### WASH : breakdown of activities, overall
```{r wash_activ, cache=TRUE}
knitr::kable(activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(activity) %>%
        summarise(num = sum(isActive)) %>% 
          arrange(desc(num))
        )
```




### WASH : by partners: summed number of unique types of activities for ALL projects.
```{r wash_partner, cache=TRUE}
uniqueActiv = activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(Awardee) %>% 
          distinct(Awardee, activity) %>% 
          summarise(numDistinctActiv = n()) %>% 
          arrange(desc(numDistinctActiv))

kable(uniqueActiv)

uniquePerProj = activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(Awardee, id) %>% 
          summarise(numDistinctActiv = sum(isActive)) %>% 
          arrange(desc(numDistinctActiv))

# Roll up to the Awardee level.
kable(uniquePerProj %>% 
        group_by(Awardee) %>%
        summarise(avgActiv = round(mean(numDistinctActiv),1),
                  minActiv = min(numDistinctActiv),
                  maxActiv = max(numDistinctActiv),
                  sdActiv = round(sd(numDistinctActiv),2),
                  numProj = n()) %>% 
        arrange(desc(avgActiv))
)

```

### WASH : by partners: average number of WASH activities per project.
```{r wash_partner_avg, cache=TRUE}

uniquePerProj = activ %>% filter(str_detect(category1, 'WASH') | 
                         str_detect(category2, 'WASH')) %>% 
        group_by(Awardee, id) %>% 
          summarise(numDistinctActiv = sum(isActive)) %>% 
          arrange(desc(numDistinctActiv))

# Roll up to the Awardee level.
kable(uniquePerProj %>% 
        group_by(Awardee) %>%
        summarise(avgActiv = round(mean(numDistinctActiv),1),
                  minActiv = min(numDistinctActiv),
                  maxActiv = max(numDistinctActiv),
                  sdActiv = round(sd(numDistinctActiv),2),
                  numProj = n()) %>% 
        arrange(desc(avgActiv))
)

```


## What's the correlation b/w WASH activities?

```{r washActiv, cache = TRUE}
wash = activities %>% 
  filter(str_detect(category1, 'WASH') | str_detect(category2, 'WASH')) %>% 
  select(activity, isActive, id)

ggplot(wash, aes(x = y = isActive))

corMatrix = wash %>% 
  mutate(isActive = ifelse(is.na(isActive), 0, isActive)) %>% 
  spread(activity, isActive)

x = cor(corMatrix %>% select(-id))

```

